It Is Unbearable To Breath Here': Air Quality, Open Incineration, And Misinformation In Blantyre, Malawi

### 2023-05-19 manuscript/manuscript.qmd
Fix citation

Queen's, simply, produces an immense amount of waste that it is unable
to efficiently manage. The dozens of wards, offices, and kitchens,
serving the hundreds of patients and staff, generates considerable
infectious medical waste, mixed domestic and office waste, and kitchen
waste, which needs to be removed and disposed of daily. Some individual
wards implement waste separation, binning medical wastes separately from
general waste, however separation ends at the point of collection.
Queen's is not served by regular municipal waste collection. What
municipal collection that is done is ad-hoc, sporadic, and based on the
hospital's available financial resources. Rather, most hospital waste
(infectious and non-infectious) is gathered by grounds staff and burnt
at the hospital's incinerator, located at a central point within the
hospital campus[^1] [@kalina2020this]. However, the vast majority of the
deposited waste is not incinerated, as a new incinerator provided by the
Ministry of Health in late 2019, is only able to handle a small
percentage of the hospital's waste. Furthermore, the incinerator is
frequently inoperable, under repair, or without power[^2]. Rather, the
waste is openly burned, in a constantly smouldering pile to the side of
the incinerator building, sending up the aforementioned clouds of smoke,
which choke the grounds (@fig-open-burning). Innumerable other, smaller,
fires, scattered across the QECH campus (@fig-map-sensors) contribute
their part, as grounds and maintenance staff habitually burn leaves and
other garden refuse. Furthermore, caregivers[^3] who reside separately
on hospital grounds, cook for themselves throughout the day and night
using biomass (charcoal or wood), and burn their own domestic waste. As
a result, air quality at the hospital is a constant source of
discussion, with patients, staff, and caregivers struggling to cope,
when just breathing is, as the quote[^4] used in the title of this
article, 'unbearable.'

@@

[^4]: ID5, 2019-11-18

Challenges with solid waste management (SWM) are not unique to Queen's,
and remain persistent globally, particularly within African nations, and
the Global South more broadly [@bundhoo2018solid;
@sergekubanza2022exploring; @muheirwe2022paradox; @srivastava2015urban].
Explanations for these disparities have centred on the on waste
generation outstripping waste management infrastructure
[@muheirwe2022paradox], and the high costs of waste management systems,
which often prove unaffordable for many low-and-middle-income countries
(LMICs) [@guerrero2013solid; @sergekubanza2022exploring]. Waste
collection is an important step within SWM systems, and is a common
barrier within LMIC contexts [@bundhoo2018solid]. Waste inequalities
between high-income countries (HICs) and low-income countries (LICs) is
often most pernicious and visible at the point of collection, with some
HICs achieving 100% collection rates, while some African nations have
rates well below 50% [@shi2021unbalanced]. Inequalities persist beyond
collection, however. Even once waste is collected, municipalities in
LMICs may burn or dump it in non-sanitary landfills due to the absence
of further treatment or disposal options [@cook2021globala]. According
to @kaza2018what, 90% of waste in LIC is disposed of in unregulated
dumps or openly burned, while the quantity of waste generation in these
countries is expected to triple by 2050. As a result, open waste
burning, in nations like Malawi, can be expected to become more
prevalent.

>>>>

Queen's, simply, produces an immense amount of waste that it is unable
to efficiently manage. The dozens of wards, offices, and kitchens,
serving the hundreds of patients and staff, generates considerable
infectious medical waste, mixed domestic and office waste, and kitchen
waste, which needs to be removed and disposed of daily. Some individual
wards implement waste separation, binning medical wastes separately from
general waste, however separation ends at the point of collection.
Queen's is not served by regular municipal waste collection. What
municipal collection that is done is ad-hoc, sporadic, and based on the
hospital's available financial resources. Rather, most hospital waste
(infectious and non-infectious) is gathered by grounds staff and burnt
at the hospital's incinerator, located at a central point within the
hospital campus[^1] [@tilley2020wea]. However, the vast majority of the
deposited waste is not incinerated, as a new incinerator provided by the
Ministry of Health in late 2019, is only able to handle a small
percentage of the hospital's waste. Furthermore, the incinerator is
frequently inoperable, under repair, or without power[^2]. Rather, the
waste is openly burned, in a constantly smouldering pile to the side of
the incinerator building, sending up the aforementioned clouds of smoke,
which choke the grounds (@fig-open-burning). Innumerable other, smaller,
fires, scattered across the QECH campus (@fig-map-sensors) contribute
their part, as grounds and maintenance staff habitually burn leaves and
other garden refuse. Furthermore, caregivers[^3] who reside separately
on hospital grounds, cook for themselves throughout the day and night
using biomass (charcoal or wood), and burn their own domestic waste. As
a result, air quality at the hospital is a constant source of
discussion, with patients, staff, and caregivers struggling to cope,
when just breathing is, as the quote[^4] used in the title of this
article, 'unbearable.'

@@

[^4]: ID5, 2019-11-18
x
Challenges with solid waste management (SWM) are not unique to Queen's,
and remain persistent globally, particularly within African nations, and
the Global South more broadly [@bundhoo2018solid;
@sergekubanza2022exploring; @muheirwe2022paradox; @srivastava2015urban].
Explanations for these disparities have centred on the on waste
generation outstripping waste management infrastructure
[@muheirwe2022paradox], and the high costs of waste management systems,
which often prove unaffordable for many low-and-middle-income countries
(LMICs) [@guerrero2013solid; @sergekubanza2022exploring]. Waste
collection is an important step within SWM systems, and is a common
barrier within LMIC contexts [@bundhoo2018solid]. Waste inequalities
between high-income countries (HICs) and low-income countries (LICs) is
often most pernicious and visible at the point of collection, with some
HICs achieving 100% collection rates, while some African nations have
rates well below 50% [@shi2021unbalanced]. Inequalities persist beyond
collection, however. Even once waste is collected, municipalities in
LMICs may burn or dump it in non-sanitary landfills due to the absence
of further treatment or disposal options [@cook2021globala]. According
to @kaza2018what, 90% of waste in LIC is disposed of in unregulated
dumps or openly burned, while the quantity of waste generation in these
countries is expected to triple by 2050. As a result, open waste
burning, in nations like Malawi, can be expected to become more
prevalent.
